Exploring colonial impacts on indigenous hunting and animal use in northeastern Taiwan

### 2025-03-03 analysis/paper/paper.qmd
work on discussion section

---
title: "Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest Iron Age settlement in northeastern Taiwan [@Chen2007]. Kiwulan encountered European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. What was the resource use of Indigenous people before and after the foreign presence? Can we identify shifts in these practices in response to colonial economic demands? To address these questions, I will focus on subsistence practices because the choices made in daily practices can reveal the strategies Indigenous people used to respond to resource explortations by foreign presence. I will examine the zooarchaeological assemblages from the upper culture layer at Kiwulan, a large settlement dating to 14-19th centuries. Zooarchaeology can offer insights into how animals were utilized in response to economic strategies within certain social contexts, such as European colonization or intercultural contact [@Defrance2009]. Through a zooarchaeological perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

@@

```{r}
#| label: taxa-NISP-counts
#| echo: false
# taxa counts
fauna_combined_taxa <-
  fauna_combined_context_all %>%
  count(taxa, `Weight (g)`) %>%
  mutate(taxa = str_replace(taxa, "cervidae", "cervid")) %>%
  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Indeterminate", tolower(taxa), sep =" "),
                          TRUE ~ taxa)) %>%
  mutate(`Common name` = case_when(str_detect(taxa, "Cer") ~ "Sika deer", str_detect(taxa, "Bub") ~ 
                                   "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",
                                   str_detect(taxa, "Ree") ~ "Muntjac", taxa == "Sus scrofa" ~ "Wild boar",
                                   str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~ 
                                   "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~ 
                                   "Ray-finned fish", str_detect(taxa, "Tes") ~ "Turtle",
                                   str_detect(taxa, "cervid") ~ "Deer", taxa == "Sus scrofa domestica" ~ 
                                   "Domestic pig", taxa == "Aves" ~ "Birds")) %>%
  mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
                           str_detect(taxa, "Sus") ~ 3, str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 4,
                           str_detect(taxa, "munt") ~ 5, str_detect(taxa, "Rat") ~ 6, 
                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Tes") ~ 8,
                           str_detect(taxa, "Ave") ~ 9, str_detect(taxa, "large m") ~ 10,
                           str_detect(taxa, "medium m") ~ 11, TRUE ~ 12)) %>%
  group_by(group) %>%
  arrange(desc(`Weight (g)`), .by_group = TRUE) %>%
  ungroup () %>%
  rename(Taxon = taxa, NISP = n) %>%
  select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
  bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>% 
  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`))

@@

The faunal assemblage analyszed in this paper from Kiwulan is composed of `r total_NISP` bone fragments based on NISP, which increased over time from CL6, peaked at CL4, and then declined dramatically after CL4 (@fig-NISP-proportion: A). A similar trend was also observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B). Among the faunal remains, the deer family is the majority accounting for about `r deer_fami_NISP_per`% of the bones identified to class level, followed by suid and cattle as the second and third taxon (@tbl-counts). Within the deer family, sika deer is prevalent, with a few bones identified as muntjac and a specimen indicating a possible affiliation with sambar deer. Although sika deer was the major animal source at the site, variations in their quantities and proportions could be observed between temporal sequences across 600 years. To further compare general patterns of dietary practices in terms of their economic values over time, the deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle. Similarly, wild boar and domestic pig (n = 2, from CL4 and CL6) were grouped into the category of suid. Rats (Rattus sp.) were excluded from further analyses as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.

@@

Burned bones (@fig-cutmark-img: d), characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4. Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percurssion scars [@Haynes1983; @Gifford2018; @Fisher1995], were observed frequently on long bones across different layers (@fig-fracture-img). These fractures indicate intentional marrow extraction at Kiwulan. CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5. Chewing marks were observed on a few bones that show a series of superficial elongated scores and circular depressions (@fig-fracture-img: e)[@Bello2023], suggesting carnivore gnawing. In addition, natural alterations to bones resulting from post-deposition impacts were also observed , such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@fig-fracture-img: c,d; @tbl-deer-modify).

@@

# Discussion
<!--results and chronology -->
The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption since CL6 until CL2, and then shifted to cattle in CL1. Despite deer had been a substantial part of dietary practices at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals shows changes that could reflect the indigenous reactions to foreign presence. Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences relative to foreign presences for the following discussion. The sequences include the pre-European period (CL6, CL5), the European period (CL4), post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1). It should be noted that the European period does not necessary means a solid European colonial control over the Kiwulan, but a broad impacts from European occupation and economic activities on indigenous societies in Yilan, such as getting involvement into the long-distance trade network, being part of the tribute system, or experiences of occasional European attacks.

<!--summary of overall observations-->
Faunal remains recovered from the layer associated with the European presence reveal the highest quantity, followed by the pre-European period. During the European period, in addition to the predominance of deer, an increased proportion of suid among faunal remains suggests a slightly greater reliance on suid as a secondary meat source compared to other time periods, along with a broader range of medium and small animals, such as muntjac, birds, fish, and reptiles. A similar pattern was observed in the ubiquity of animals, which shows the highest ubiquity of deer, suid, and munjact remains during the European period. Interestingly, deer bones from the pre-European period, specifically CL5, presented the same ubiquity as in the European period but in only half the quantity. This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but consumption of deer was intensified during the European period. During the Chinese period, the low quantity of deer in CL2 and the predominate of cattle in CL1 suggest a sudden shift in dietary practices after the Chinese arrival, indicating a Chinese influence and the decline of Indigenous population in this region.

<!--key results of skeletal part frequencies and butchery patterns-->
The skeletal part frequencies and butchery patterns of deer demonstrate more complete carcasses processed within the settlement during the European period and pre-European period compared to the post-European period and Chinese period. Although a more complete and evenly distributed deer parts with cut marks were found in both the European period and before European arrival, the substantial rise in head parts (e.g., crania and mandibles) and overall high quantity of bones in the European period suggests more deer were brought back intact to the settlement. The low frequency of head parts in the earliest layer and Chinese period might indicate deer could be mostly processed at the hunting location with only meaty (e.g., upper limb) and mid-meaty portions (e.g., ribs) selectively butchered and transported back. Interestingly, among the meaty portions throughout all periods, front quarters (e.g., humeri, radius) are more frequently found compared to forequarters (e.g., femur, tibia). One likely explanation is that femur and tibia might be heavily fragmented due to being broken for their high marrow yields [@Madrigal2002], leading to underestimation due to difficult identification. The behavior of marrow extraction can be supported by around one-third long bones being fractured and a moderate degree of fragmentation on average at Kiwulan. Another possibility is forequarters could be traded since deer legs as commodity were mentioned in European records [@Blusse2006, pp.220-224], but unfortunately, there was no detail about which portion of legs were normally traded.

<!--skinning-->
The cut marks found on the lower limbs of deer at Kiwulan should be associated with the traces left by skinning processes, which confirms that deer skins were removed and likely used for trading with external traders. A large quantity of foreign goods, such as ornaments, porcelains, and metal items, supports frequent trading activities between Kiwulan residents and outsiders [@Wang2020ornament; @Chen2007]. Deer hides and venison, were important commodities in the East Asian trade network since the mid-15th century, connecting Han Chinese, Japanese, and Indigenous people, especially in western Taiwan [@Cheng2017; @Hollmann1991]. This deer trade was later monopolized by the Europeans in the 17th century, and deer products were largely exported to Japan. The evidence of the removal of deer skins at Kiwulan suggests that not only western Taiwan but also northeastern Taiwan involved and impacted by the long-distance trade. Deer bones with skinning cuts did not show a substantial increase in the European period; however, the disproportionate increase in head parts hints at a likelihood of brain tanning when preparing hides. Brain tanning is a traditional method for softening and preserving hides commonly used in many Indigenous cultures, such as North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016]. The present-day Bunun group in the central highlands of Taiwan has revived hide tanning based on oral history, primarily using peanuts or pig lard, with a few examples using animal brains [@Li1999, pp. 48; @Xu1994]. Although materials for tanning used by the Kavalan people are unknown due to the loss of traditions, brain tanning is a possible method given the availability of animal brains from hunting and their high-fat content.

<!--decline of deer, ceramics, and population dynamics of Kiwulan-->
The dramatic decrease in deer bones during the Chinese period may indicate overhunting that persisted for centuries or a decline in the Indigenous population (Zhan, 1995). The double increase in suid after the European arrival may suggest an increased reliance on pigs. This implies that pigs played a more significant role in the diet after Europeans arrived. Eventually, cows replaced pigs during the Chinese period, which might be a Chinese influence.

<!--environmental changes in general and taphonomic issues-->

<!--resilence-->
By contextralizing the sampled faunal remains from the major residential area at Kiwulan   into the time periods with foreign contact, we can observe the

played a consistent role in subsistence strategies.
A taxon can have high NISP but low ubiquity if concentrated in a few contexts (e.g., feasting deposits).
A taxon with high ubiquity but low NISP may have been used frequently but in small amounts.

frequencies of deer bone elements are consistent with the proportion of joints. 
Other than quantity, the measure of presence and absence of a given taxon across the sampeld units could give insights into the frequency and distribution consistency of taxa at a settlement scale.

High-ubiquity taxa (e.g., deer, pigs) likely played a consistent role in subsistence strategies. Low-ubiquity taxa (e.g., rare birds, marine species) may indicate opportunistic hunting, ritual use, or seasonal availability. Changes in ubiquity across layers may reflect environmental shifts or human adaptations. For example, a decline in large mammal ubiquity might suggest deforestation or overhunting, leading to a shift toward smaller prey.

Issue of Taphonomic & Preservation Bias: Some taxa may be less ubiquitous due to poor bone preservation rather than actual scarcity in the past. Small and fragile bones (e.g., fish, rodents) might have low ubiquity due to taphonomic processes.

These post-depositional impacts might have led to an underestimation of cut marks due to the diffculty of surface identification.

The European maritime cultural exchange to some extent altered the material culture of the Indigenous peoples in northeastern Taiwan by adopting foreign goods or customs into their daily lives. However, a sign of indigenous resilience in the face of European influences and resource exploitation may be seen in the continuing tradition of pottery production and deer consumption with few changes, even though changes were observed through the uneven distribution of prestige items, such as glass beads and porcelains, across different households in the living area and burials (Wang and Marwick, 2020a; 2021).

The continuous practices of ceramic making and deer consumption by the Indigenous peoples before and after the European arrival can be viewed as a way of negotiating their identities and maintaining their traditions. Everyday practices are essential in connecting history and social memories, which helps maintain tradition in a setting of frequent cultural exchange or asymmetrical relations (Silliman, 2009). Moreover, ceramics could be an item that conveys social meaning through its style, such as decorations, and be used to show group identity (Croucher and Wynne-Jones, 2006; Pikirayi, 2007). It is worth noting that the pottery found at Kiwulan, with its typical round body and incised geometric motifs, is also commonly found across many sites during this period in northeastern Taiwan (Taiwan History, 2005). Similar geometric motifs and vessel shapes in the region might suggest the communication of Indigenous identity within the context of ongoing cross-cultural interactions with outsiders, including Europeans and Chinese. Furthermore, despite the availability of foreign items, such as iron pans that could have potentially replaced local pottery for cooking purposes, local ceramics have been used continuously for over 6,000 years (Chen, 2007). This emphasizes the significant role of pottery-making in the lives of Indigenous communities.

@@

Taiwan has played an important role in exploring the effect of colonial activities on Indigenous communities and local environment since the protohistoric period. This project will explore the environmental effect of colonial presence on local ecology on a regional scale using archaeological evidence that has not yet been fully understood. Archaeological materials can offer more direct evidence, allowing better interpretation for the environments where Indigenous live in that can also reduce biases from historical records. Through exploring the long-term dynamics of human-environmental relationship in colonial settings, we can better understand changes in local environments and Indigenous resilience. In addition to the Europeans, the interaction between Chinese traders and Indigenous groups in Taiwan since the Neolithic period, and an influx of Han Chinese immigrants in the 18th century, has contributed to the significant local cultural and environmental transformation. Those historical events are closely related to the Taiwan Indigenous people today, which is crucial to understanding the contemporary Indigenous cultures and socio-ecological landscape. 
The results of this project can advance our knowledge of the reaction of Indigenous people in Taiwan when facing dominant foreign power, and generate new anthropological insights about the adaptation and choices of Indigenous people. In the long term, this research will also benefit local Indigenous people, Kavalan, in northeastern Taiwan and help them to trace back their past and revitalize their cultural heritage by the plans of public presentation in the future. In addition, this project will combine geoarchaeological, geochemical, and computational methods to analyze archaeological evidence. The multidisciplinary approach can not only advance the development and application of scientific techniques in Taiwan archaeology, but also provide an opportunity to form collaborative relationships with scholars with different backgrounds, which will help to develop the academic network and create interdisciplinary collaboration.

>>>>

---
title: "Exploring colonial impacts on indigenous hunting and animal use in northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest Iron Age settlement in northeastern Taiwan [@Chen2007]. Kiwulan encountered European colonial power in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. What was the resource use of Indigenous people before and after the foreign presence? Can we identify subsistence strategies in response to impacts resulting from integration into trading network? To address these questions, I will analyze the zooarchaeological assemblages from Kiwulan, because they provide valuable insights into subsistence practices and accommodation within specific social contexts, such as European colonization or being part of the regional trade network [@Defrance2009; @Crabtree1990]. Through a perspective from subsistence, this study aims to shed light on Indigenous persistence and resilience in the face of foreign presence.

@@

```{r}
#| label: taxa-NISP-counts
#| echo: false
# taxa counts
fauna_combined_taxa <-
  fauna_combined_context_all %>%
  count(taxa, `Weight (g)`) %>%
  mutate(taxa = str_replace(taxa, "cervidae", "cervid")) %>%
  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Indeterminate", tolower(taxa), sep =" "),
                          TRUE ~ taxa)) %>%
  mutate(`Common name` = case_when(str_detect(taxa, "Cer") ~ "Sika deer", str_detect(taxa, "Bub") ~ 
                                   "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",
                                   str_detect(taxa, "Ree") ~ "Muntjac", taxa == "Sus scrofa" ~ "Suid",
                                   str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~ 
                                   "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~ 
                                   "Ray-finned fish", str_detect(taxa, "Tes") ~ "Turtle",
                                   str_detect(taxa, "cervid") ~ "Deer", taxa == "Sus scrofa domestica" ~ 
                                   "Domestic pig", taxa == "Aves" ~ "Birds")) %>%
  mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
                           str_detect(taxa, "Sus") ~ 3, str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 4,
                           str_detect(taxa, "munt") ~ 5, str_detect(taxa, "Rat") ~ 6, 
                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Tes") ~ 8,
                           str_detect(taxa, "Ave") ~ 9, str_detect(taxa, "large m") ~ 10,
                           str_detect(taxa, "medium m") ~ 11, TRUE ~ 12)) %>%
  group_by(group) %>%
  arrange(desc(`Weight (g)`), .by_group = TRUE) %>%
  ungroup () %>%
  rename(Taxon = taxa, NISP = n) %>%
  select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
  bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>% 
  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`)) %>% 
  mutate(Taxon = ifelse(Taxon == "Sus scrofa", "Sus spp.", Taxon))

@@

The faunal assemblage analyszed in this paper from Kiwulan is composed of `r total_NISP` bone fragments based on NISP, which increased over time from CL6, peaked at CL4, and then declined dramatically after CL4 (@fig-NISP-proportion: A). A similar trend was also observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B). Among the faunal remains, the deer family is the majority accounting for about `r deer_fami_NISP_per`% of the bones identified to class level, followed by suid and cattle as the second and third taxon (@tbl-counts). Within the deer family, sika deer is prevalent, with a few bones identified as muntjac and a specimen indicating a possible affiliation with sambar deer. Although sika deer was the major animal source at the site, variations in their quantities and proportions could be observed between temporal sequences across 600 years. To further compare general patterns of dietary practices in terms of their economic values over time, the deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle. Similarly, domestic pig (n = 2, from CL4 and CL6) and wild boar were grouped into the category of suid. Rats (Rattus sp.) were excluded from further analyses as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.

@@

Burned bones (@fig-cutmark-img: d), characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4. Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percurssion scars [@Haynes1983; @Gifford2018; @Fisher1995], were observed frequently on long bones across different layers (@fig-fracture-img). These fractures indicate intentional marrow extraction at Kiwulan. CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5. Chewing marks were observed on a few bones that show a series of superficial elongated scores and circular depressions (@fig-fracture-img: e)[@Bello2023], suggesting carnivore gnawing. In addition, natural alterations to bones resulting from post-deposition impacts were also observed , such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@fig-fracture-img: c,d; @tbl-deer-modify). Those post-depositional impacts might have led to an underestimation of cut marks due to the difficulty of surface identification.

@@

# Discussion
<!--results and chronology -->
The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption from CL6 to CL2, and then shifted to cattle in CL1. Despite deer was a substantial part of dietary practices at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals show changes that could reflect the indigenous reactions to foreign presence. Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences in relation to foreign presences for further discussion. The temporal sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1). It should be noted that the European period does not necessarily mean a solid European colonial control over the Kiwulan, but a broad impact from European occupation in Northern Taiwan and associated colonial activities on indigenous societies in Yilan, such as getting involved in the long-distance trade network, being part of the tribute system, and experiencing occasional European attacks.

<!--summary of overall observations across time periods -->
Faunal remains recovered from the European period reveal the highest quantity, followed by the late pre-European period. During the European period, in addition to the predominance of deer, an increased proportion of suid suggests a slightly greater reliance on suid as a secondary meat source. A similar pattern was observed from a higher ubiquity of suids during the European period compared to other periods. Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but in only half the quantity. This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but deer utilization was intensified during the European period. The dramatic decrease in deer bones after the European period might be related to both the decline of the Kiwulan population and a decrease in the wild deer population. Sika deer populations in western Taiwan have significantly shrunk since the 18th century because of habitat loss and overhunting [@Chen2000animal]. Despite this, deer remained a predominant resource for local consumption at Kiwulan during the post-European and early Chinese period. In the late Chinese period, the dietary preference shifts from deer to cattle, suggesting a Chinese influence and the relocation of the Kavalan people to Hualien due to population pressures from Chinese immigration [@Zhan1995].

<!--skeletal part frequencies and butchery patterns of deer-->
Despite being the main resource throughout most periods, deer utilization shows different patterns over time. Dear body parts and butchery patterns demonstrate more complete carcasses processed within the settlement during the European and pre-European period compared to the post-European and Chinese period. Although more complete and evenly distributed joints with cut marks were found in both the European period and before the European arrival, the substantial rise in head parts (e.g., crania and mandibles) with the overall high quantity suggests more deer were brought back intact to the settlement. On the contrary, the low frequency of head parts in the early pre-European period and Chinese period might indicate deer could be mostly processed at the hunting location with mostly meaty (e.g., upper limb) and mid-meaty portions (e.g., ribs) selectively butchered and transported back. Interestingly, among the meaty portions, front quarters (e.g., humeri, radius) are more frequently found compared to forequarters (e.g., femur, tibia) throughout all periods. One likely explanation is that femur and tibia might be heavily fragmented due to being broken for their high marrow yields [@Madrigal2002], leading to underestimation due to difficult identification. Marrow extraction at Kiwulan can be supported by around one-third of long bones being fractured and a moderate degree of fragmentation on average. Another possibility is forequarters could be traded since deer legs and meats as commodities were commonly mentioned in European records [@Blusse2006, pp.220-224], but there is no mention of any specific legs commonly traded. Considering the generally high frequencies of meaty portions found at Kiwulan, deer might not just for consumption at the site, but also for trading.

<!--deer hides-->
Deer hides were important commodities since the mid-15th century, connecting Han Chinese, Japanese, and Indigenous people, and later monopolized by the Europeans in the 17th century, when deer products were largely exported to Japan, especially from western Taiwan [@Cheng2017; @Hollmann1991]. The cut marks found on the lower limbs of deer at Kiwulan should be associated with the traces left by skinning, which confirms the removalof  deer skins and likely being traded with external traders. A large quantity of foreign goods, such as ornaments, porcelains, and metal items, supports frequent exchange between Kiwulan residents and outsiders [@Wang2020ornament; @Chen2007]. The evidence of skinning and foreign goods at Kiwulan suggests that not only western Taiwan but also northeastern Taiwan was involved and impacted by the long-distance trade. Although deer bones with skinning cuts did not show a substantial increase in the European period, the disproportionate increase in head parts might hint at a likelihood of brain tanning when preparing hides. Brain tanning is a traditional method for softening and preserving hides commonly used in many Indigenous cultures, such as North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016]. The present-day Bunun group in the central highlands of Taiwan has revived hide tanning based on oral history, primarily using peanuts or pig lard, with a few examples using animal brains [@Li1999, pp. 48; @Xu1994]. Despite the materials for tanning used by the Kavalan people being unknown due to the loss of traditions, brain tanning is a practicable method given the availability of deer brains from hunting.

<!-- population dynamics and arguments-->
The population dynamics at Kiwulan must be considered when discussing faunal changes over time. A larger pottery vessels during the European period suggests population growth [@Wang2020pot; @Hsieh2009], which may contribute to the increasing animal consumption overall. However, the impacts from foreign presence on faunal assemblages can be revealed through butchery patterns of deer parts, the relative proportions of animals, and taxon ubiquity. First, deer bones with skinning evidence were mostly found during the European period, followed by the late pre-European phase, corresponding with the rise of regional trade networks since the 15th century and intensified in the 17th century. Second, if the high frequency of deer head parts during the European period solely resulted from food scarcity due to population growth and increased consumption of brains for nutrition, a parallel rise in marrow extraction would be expected. However, fractured long bones did not increase during the European period or the previous phase; instead, they were 10-30 % lower than in other periods. Third, population growth should have led to a more proportional increase in both large and small animals. Although there was a high taxonomic diversity during the European period, suids present a disproportionate rise in their ubiquity, quantity, and proportion compared to other animals. This suggests that intensified deer hunting for trade may have driven indigenous people to rely more on suids as an alternative prey, likely due to their comparable body size and meat return rates to deer [@Broughton2011prey]. Suids as a secondary protein source after deer is also observed in prehistoric western Taiwan [@Chuang2021].

<!--indigenous reaction and resilience-->
Compelling the results of faunal analysis and reports from historical records, the substantial increase in deer bones in the European period not just simply the result of population growth, but also a shift driven by European presence and associated intensified long-distance trade. It is clear to see deer is both the first priority of food source and commodities at Kiwulan since the late pre-European period and intensified during the European period. Based on the dietary preference, the increased use of suids during the European period could be a temporary subsistence strategy in response to the high demands of deer products. After the Europeans left, deer hide trade continued but on a smaller scale, with around 10,000 hides exported annually in the 18th century, which is 20 times less than during the European period [@]. This change likely reduced the reliance of suids, as shown by their decreased proportion at Kiwulan. The slightly increase in muntjact during the post-European and early Chinese period might be associated with the decline of deer population, leading people turned to use muntjact hides for trade instead []. Nevertheless, deer continued to be consumed until the early Chinese period. Everyday practices play an essential role for connecting history and social memories, enabling Indigenous people maintain tradition in a setting of frequent cultural exchange or asymmetrical relations [@Silliman2009; e.g., @Noe2023]. The persistence of deer consumption at Kiwulan reflect identity negotiation and tradition maintenance, while the increased utilization of suids likely responded to the high demand for deer products in large scale trade. This highlights how Kiwulan residents reacted to foreign presence and demonstrated cultural resilience.

<!--taphonomic & preservation bias -->

@@

Taiwan has played an important role in exploring the effect of colonial activities on Indigenous communities and local environment since the protohistoric period. This project will explore the environmental effect of colonial presence on local ecology on a regional scale using archaeological evidence that has not yet been fully understood. Archaeological materials can offer more direct evidence, allowing better interpretation for the environments where Indigenous live in that can also reduce biases from historical records. Through exploring the long-term dynamics of human-environmental relationship in colonial settings, we can better understand changes in local environments and Indigenous resilience. In addition to the Europeans, the interaction between Chinese traders and Indigenous groups in Taiwan since the Neolithic period, and an influx of Han Chinese immigrants in the 18th century, has contributed to the significant local cultural and environmental transformation. Those historical events are closely related to the Taiwan Indigenous people today, which is crucial to understanding the contemporary Indigenous cultures and socio-ecological landscape.

The results of this project can advance our knowledge of the reaction of Indigenous people in Taiwan when facing dominant foreign power, and generate new anthropological insights about the adaptation and choices of Indigenous people. In the long term, this research will also benefit local Indigenous people, Kavalan, in northeastern Taiwan and help them to trace back their past and revitalize their cultural heritage by the plans of public presentation in the future. In addition, this project will combine geoarchaeological, geochemical, and computational methods to analyze archaeological evidence. The multidisciplinary approach can not only advance the development and application of scientific techniques in Taiwan archaeology, but also provide an opportunity to form collaborative relationships with scholars with different backgrounds, which will help to develop the academic network and create interdisciplinary collaboration.
